---
title: "2,3-DPG and pH in stored RBC"
subtitle: "Reproducible analysis — *Journal of Neurotrauma* submission"
author:
  - name: Juan Sahuquillo
  - name: Marta Peris
date: "2026-02-02"
date-modified: "2026-06-18"
abstract: |
  The efficacy of red blood cell (RBC) transfusions depends on oxygen
  delivery, which deteriorates during storage due to the "storage lesion,"
  notably 2,3-diphosphoglycerate (2,3-DPG) depletion. We characterized
  the temporal dynamics of 2,3-DPG depletion and metabolic acidification
  in 24 leukodepleted packed RBC units stored under standard conditions
  from day 3 to 42. 2,3-DPG concentrations (µmol/g Hb) were quantified
  using liquid chromatography–mass spectrometry (LC–MS) with isotopically
  labeled internal standards, alongside parallel pH measurements. Baseline
  (day 3) 2,3-DPG levels showed substantial inter-unit variability (median
  4.8 µmol/g Hb; range 0.8–14.7). A rapid decline occurred across all
  units, dropping to ~27% of baseline by day 7, >90% by day 14, and
  falling below the detection limit after day 17. Mixed-effects modeling
  confirmed a strong negative association between storage duration and
  2,3-DPG (β = −0.298 µmol/g Hb/day; p < 0.001). Concurrently, pH
  decreased progressively from 7.06 ± 0.06 to 6.40 ± 0.04 (p < 0.001).
  In conclusion, 2,3-DPG depletion is an early, rapid phenomenon linked
  to acidification that impairs oxygen unloading. Preserving 2,3-DPG is
  a critical, modifiable target for improving stored RBC quality,
  suggesting that storage duration alone may be an inadequate surrogate
  for RBC metabolic quality.
editor: visual
editor_options:
  chunk_output_type: console
format:
  html:
    toc: true
    number-sections: true
    self-contained: true
---

# Settings for R session

```{r}
#| label: setup
#| message: false
#| warning: false

Sys.setenv(LANG = "en")
graphics.off()
options(scipen=999, digits=4)
```

# Load libraries

```{r}
#| label: packages
#| message: false
#| warning: false
library(broom.mixed) # for tidying up linear models 
library(emmeans) # for the joint_tests() function 
library(flextable)
library(ggthemes)
library(here)
library(janitor) # cleaning column names
library(kableExtra) #nice tles
library(lme4)
library(lmerTest)
library(patchwork)
library(psych)
library(stargazer)
library(sjPlot)
library(tidyverse)

```

# Import 2,3_DPG dataset

```{r}
# Establish the file path
path <- here("data", "2_3_DPG.rds")
path
mydata <- readRDS(path)

glimpse(mydata)
```

# Convert Day 3 to 3 into a new column (day_num)

```{r}
mydata <- mydata |> 
  dplyr::mutate(day_num = readr::parse_number(day))|>
  dplyr::select(day_num, everything())
head(mydata)
```

# Pivot longer

```{r}
#| label: pivot-long
long_data<- mydata |> 
  pivot_longer(
    cols = -c(day_num, day),
    names_to = "bag_id",
    values_to = "dpg_level"
  ) |> 
  filter(!is.na(dpg_level))
head(long_data)
```

# Summarize by day

```{r}
#| label: descriptive-stats-dpg

library(psych)
desc_dpg <- psych::describeBy(long_data$dpg_level, group = long_data$day_num, mat = TRUE)
desc_dpg

```

# 4. Calculate Summary Statistics

```{r}
long_summary <- long_data |> 
  group_by(day_num) |> 
  summarize(
    N       = n(),
    Mean    = mean(dpg_level, na.rm = TRUE),   # Added na.rm = TRUE for safety
    Median  = median(dpg_level, na.rm = TRUE),
    SD      = sd(dpg_level, na.rm = TRUE),
    min     = min(dpg_level, na.rm = TRUE),
    max     = max(dpg_level, na.rm = TRUE)
  ) |> 
  # 1. Explicitly sort by time step to guarantee row 1 is the true baseline
  arrange(day_num) |> 
  # 2. Calculate percentage relative to the true baseline day
  mutate(Pct_left = (Mean / first(Mean)) * 100)

print(long_summary)

```

# Check for normality

```{r}
day3_dpg <- long_data |> 
  filter(day_num == 3) |> 
  pull(dpg_level)

summary(day3_dpg)
hist(day3_dpg, main = "Day 3 DPG Levels", col = "lightblue")
qqnorm(day3_dpg)
qqline(day3_dpg, col = "red")
shapiro.test(day3_dpg)
```

# Boxplot 2,3-DPG by day

```{r}
plot1<-ggplot(long_data, aes(x = factor(day_num), y = dpg_level)) +
  stat_boxplot(geom = "errorbar", width = 0.3, color = "gray30", size = 0.62) +
  geom_boxplot(fill = "gray90", color = "gray30", outlier.alpha = 0.5, linewidth = 0.62) +
  geom_jitter(size = 3, width = 0.1, alpha = 0.65, color = "steelblue") +
    theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
    axis.title.x = element_text(face = "bold", size = 13),
    axis.title.y = element_text(face = "bold", size = 13),
    axis.text.x = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 12)
  ) +
  labs(
    title = "2,3-DPG by Day",
    subtitle = "",
    x = "Storage Day",
    y = "2,3-DPG (µmol/g Hb)"
  )
plot1
```

# Spaghetti plot

```{r}
plot2<-ggplot(long_data, aes(x = day_num, y = dpg_level, group = bag_id)) +
  geom_line(color = "steelblue", alpha = 0.5, size = 0.7) +
  geom_point(color = "steelblue", alpha = 0.6, size = 2) +
  
  # 3. Add a heavy black line for the Population Mean
  stat_summary(aes(group = 1), fun = mean, geom = "line", 
               color = "red", size = 1.2) +
theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    axis.title = element_text(face = "bold", size = 14),
    axis.text = element_text(face = "bold", size = 12)
  ) +
  labs(
    title = "Individual 2,3-DPG trajectories",
    subtitle = "",
    x = "Storage Day",
    y = "2,3-DPG (µmol/g Hb)"
  ) +
  # X-axis only shows the days you measured
  scale_x_continuous(breaks = c(3, 7, 10, 14, 17))
plot2

```

# Join plots

```{r}
library(patchwork)
plot1 + plot2

```

# Linear mixed models analysis for 2,3-DPG

```{r}
#| label: model1-dpg-lmm
#| message: false
#| warning: false

# LMM: 2,3-DPG ~ storage day with random intercept for pRBC unit
# REML = TRUE for unbiased variance component estimation
model1 <- lmer(dpg_level ~ day_num + (1 | bag_id), data = long_data, REML = TRUE)
summary(model1)

# Note: lmer() returns a singular fit (random intercept variance = 0),
# indicating no detectable between-bag baseline heterogeneity after
# accounting for the strong time trend. Cluster-robust SEs (CR2) via
# clubSandwich are reported to account for the repeated-measures structure.
robust_test <- clubSandwich::coef_test(
  model1,
  vcov       = "CR2",
  cluster    = long_data$bag_id,
  test       = "Satterthwaite"
)
robust_test
```

# pH analysis

```{r}
path_ph <- here("data", "Stored_blood_pH.rds")
path_ph
mydata_ph <- readRDS(path_ph)
head(mydata_ph)
```

# Convert Day 3 to 3

```{r}
mydata_ph <-mydata_ph |> 
  mutate(day_num = readr::parse_number(day))|> 
  select(day_num, everything())

mydata_ph$day_num <- factor(
  mydata_ph$day_num,
  levels = c(3,7,11,14,17,21,25,28,32,37,42)
)
head(mydata_ph)
```

#Pivot longer the pH dataframe

```{r}
long_data_ph<- mydata_ph |> 
  pivot_longer(
    cols = -c(day_num, day),
    names_to = "bag_id",
    values_to = "pH"
  ) |> 
  filter(!is.na(pH))
head(long_data_ph)
```

# Summarize pH by day

```{r}

library(psych)
psych::describeBy(long_data_ph$pH, group = long_data_ph$day_num, mat = TRUE)

```

# Calculate Summary Statistics

```{r}
long_summary_ph <- long_data_ph |>
  group_by(day_num) |> 
  summarize(
    N = n(),
    Mean = mean(pH, na.rm = TRUE),
    Median = median(pH, na.rm = TRUE),
    SD = sd(pH, na.rm = TRUE),
    min = min(pH, na.rm = TRUE),
    max = max(pH, na.rm = TRUE)
  ) |> 
  mutate(Pct_left = (Mean / first(Mean)) * 100)

print(long_summary_ph)
```

# Boxplot pH by day

```{r}
plot3 <- ggplot(long_data_ph, aes(x = factor(day_num), y = pH)) +
stat_boxplot(geom = "errorbar", width = 0.3, height = 0.2, color = "gray30", size = 0.62) +
geom_boxplot(fill = "orange", color = "gray30", outlier.alpha = 0.5, linewidth = 0.62) +
coord_cartesian(ylim = c(6.25, 7.40)) +
scale_y_continuous(breaks = seq(6.25, 7.40, 0.1))+
geom_jitter(size = 2, width = 0.1, alpha = 0.65, color = "steelblue") +
theme_minimal(base_size = 12) +
theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 13),
    axis.title.x = element_text(face = "bold", size = 13),
    axis.title.y = element_text(face = "bold", size = 13),
    axis.text.x = element_text(face = "bold", size = 12,),
    axis.text.y = element_text(face = "bold", size = 12)
  ) +
labs(
  title = "",
  subtitle = "",
  x = "Storage Day",
  y = "pH"
  )

plot3
```

# Linear mixed models analysis for pH

```{r}
str(long_data_ph)
head(long_data_ph)
# Linear model
lin.model<- lm (formula = pH ~ day_num, data = long_data_ph)
summary(lin.model)
print(lin.model)
predicted <- predict(lin.model)

# Deconstruct lin.model
x <- as.numeric(long_data_ph$day_num)
y <- long_data_ph$pH

y.pred <- 7.0021 + x*(-0.0163)
SS.resid <- sum((y - y.pred)^2 )
SS.tot <- sum((y - mean(y))^2 )
r <- cor(x, y)
R.squared <- 1 - (SS.resid / SS.tot)
print(SS.tot)
print(SS.resid)
print(R.squared)
print( r^2 )

```

# Linear Mixed Model for pH

```{r}
library(lme4)
library(lmerTest)
library(emmeans)
library(dplyr)

# Ensure correct ordering
long_data_ph <- long_data_ph |>
  arrange(day_num) |>
  mutate(day_num = factor(day_num, levels = unique(day_num)))

# Confirm dataset
long_data_ph |> count(day_num)

# Fit model
model_ph <- lmer(
  pH ~ day_num + (1 | bag_id),
  data = long_data_ph,
  REML = TRUE
)

summary(model_ph)

# Estimated marginal means
emm_ph <- emmeans(
  model_ph,
  ~ day_num,
  lmer.df = "satterthwaite"
)

summary(emm_ph)

# Contrasts vs Day 3
selected_contrasts <- contrast(
  emm_ph,
  method = "trt.vs.ctrl",
  ref = 1
)

summary(selected_contrasts, infer = TRUE)

# Singularity check
isSingular(model_ph)

```

# sessionInfo()

```{r}
sessionInfo()
```
